---
output: html_document
---
# Basic Inferential Analysis of Data on Vitamin C and Tooth Length in Guinea Pigs
By Matthew Sedlar  

## Overview

This is a basic inferential data analysis of E.W. Crampton's 1947 study on the effect of vitamin C on the growth of the odontoblast of the incisor teeth in guinea pigs. It tests whether one delivery method or a combination of delivery method and dosage per day was more successful at odontoblast growth.  

## About the Data

The ToothGrowth data is included in R and looks at the growth of the odontoblast of the incisor teeth after vitamin C was given to guinea pigs via one of two delivery methods. The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC)).

The ToothGrowth data is provided as a data frame with 60 observations on 3 variables.

* [,1]	len	numeric	Tooth length
* [,2]	supp	factor	Supplement type (VC or OJ).
* [,3]	dose	numeric	Dose in milligrams/day

## Loading the Data

The data set comes packaged with R, so there's very little for me to do here. I will designate a new object to copy the data to, since I don't like modifying, if that's necessary, the original data.

```{r}

# calling the data set
data("ToothGrowth")

# making a copy of the data
data <- ToothGrowth

```

## Hypothesis

The first step is to conduct an exploratory analysis of the data to find any interesting patterns. I will start off with a boxplot that shows tooth growth by delivery type.

```{r echo=FALSE, message=FALSE, fig.height=3.75}
library(ggplot2)

ggplot(data) +
  geom_boxplot(aes(supp,len,fill=supp)) +
  xlab("Delivery Type") +
  ylab("Length") +
  ggtitle("Tooth Length by Delivery Type")

```

While the range of tooth length with the ascorbic acid delivery type is much more spread out, the bulk of variables -- everything between Q1 and Q3 -- for the orange juice delivery type is higher. Is orange juice more effective than ascorbic acid? What about dosage?

```{r echo=FALSE, message=FALSE, fig.height=3.75}

ggplot(data) +
  geom_point(aes(dose,len), pch=1,cex=5,alpha=.5) +
  ylab("Tooth Length") +
  xlab("Dosage") +
  ggtitle("Tooth Length by Daily Dosage")
  
```

It also looks like there is a positive correlation between a higher dosage and tooth length.

I want to test if the supplement type or a combination of supplement type and dosage of supplement have an effect on tooth growth. To do this, I have to determine my null hypothesis. Since my hypothesis is that one supplement type is more efficient than the other, my null hypothesis is that both the means of the supplement types are equal, or: 

$$H_{o}:\overline{X}=\overline{Y}$$
$$H_{1}:\overline{X}>\overline{Y}$$

## Analysis

First I'm going to look at the overall tooth growth for each supplement type. I'm going to load the dplyr package, split my observations into two separate data frames and compare the variances.

```{r message=FALSE}

library(dplyr)

oj <- data %>% filter(supp == "OJ")
vc <- data %>% filter(supp == "VC")

var(oj$len) == var(vc$len)

```

With unequal variances, I will use the Welch Two Sample t-test for my confidence testing:

$$\overline{Y} - \overline{X} \pm t_{df} \times \left( \dfrac{s^{2}_{x}}{n_{x}} +  \dfrac{s^{2}_{y}}{n_{y}} \right)^\frac{1}{2}$$ 


Or I can use the simple t.test function in R with the parameter var.equal=FALSE. It should be noted that by default R's t.test function tests using a 95% confidence interval, so that is what I will be using for this analysis.

```{r}
t.test(oj$len,vc$len,var.equal = FALSE)
```

Since the p-value is greater than 0.05, there is no significant difference between the means of my samples. I cannot reject the null hypothesis in the case of comparing supplement types alone.

How about by dosages? Plotting the lengths by dosage for each supplement type shows that lower dosages with orange juice appear to be more effective than lower doses with ascorbic acid.

```{r echo=FALSE, message=FALSE, fig.height=3.75}

ggplot(data, aes(dose,len, fill=supp)) +
  geom_histogram(stat="identity") +
  facet_grid(.~supp) +
  ggtitle("Length and Dosage Per Day by Delivery System")

```

I looked at the difference in supplement types for each dosage level (0.5 mg/day, 1.0 mg/day, and 2.0 mg/day) and ran a Welch Two Sample t-test on each pairing (see Appendix A for script and results). 

For both the 0.5 mg/day (p-value = 0.006359) and 1.0 mg/day dosages (p-value = 0.001038), there actually was a significant difference in the means, rejecting the null hypothesis. For the 2.0 mg/day dosage, there was no significant difference (p-value = 0.9639).

## Conclusion 

In the sample population, orange juice (at dosages of 0.5 and 1.0 mg/day) was a better delivery system for growing guinea pig teeth than ascorbic acid. However, with this conclusion I had to make the assumption that the data is unbiased. For example, there's an assumption that the guinea pigs were not receiving extra vitamin C through their diet and that both sample groups were receiving the same diet.

As always, with studies like these where there is such an incredibly small sample size, further research is required.  

\pagebreak

# Appendix

## Appendix A

### T-tests for Each Dosage by Supplement

```{r}

# 0.5 dosage test
t.test(data[data$supp=="OJ" & data$dose==0.5,]$len,
       data[data$supp=="VC" & data$dose==0.5,]$len,
       var.equal = FALSE)

# 1.0 dosage test
t.test(data[data$supp=="OJ" & data$dose==1.0,]$len,
       data[data$supp=="VC" & data$dose==1.0,]$len,
       var.equal = FALSE)

# 2.0 dosage test
t.test(data[data$supp=="OJ" & data$dose==2.0,]$len,
       data[data$supp=="VC" & data$dose==2.0,]$len,
       var.equal = FALSE)

```

