# Statistical Inference Course Project: Basic Inferential Data Analysis
By Matthew Sedlar  

## Overview

## About the Data

The ToothGrowth data looks at the growth of the odontoblast of the incisor teeth after vitamin C was given to guinea pigs via one of two delivery methods. The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

The ToothGrowth data is provided as a data frame with 60 observations on 3 variables.

* [,1]	len	numeric	Tooth length
* [,2]	supp	factor	Supplement type (VC or OJ).
* [,3]	dose	numeric	Dose in milligrams/day

## Loading the Data

```{r}
data("ToothGrowth")
```

## Hypothesis

I want to test if the supplement type, dosage or a combination of both have an effect on tooth growth. To do this, I have to determine my null hypothesis. Since I don't have access to the average growth of guinea pig teeth (who does?!?), I've determined the best measure is the average length of teeth over all subjects. That way, if I single out one portion of the sample and it performs better than the average, I know there's something going on there.

The average growth is `r round(mean(ToothGrowth$len),2)`, which makes my null hypothesis:

$$H_{o}:\mu=18.81$$

Since my standard deviation is `r round(sd(ToothGrowth$len),2)`, I have calculated the standard error of the mean to be `r round(sd(ToothGrowth$len)/sqrt(60),2)`. I want to choose a constant so that the probability of my sample mean being greater than the constant is 5%, so I will use the following formula:

```{r echo=FALSE}

constant <- 18.81+.099*1.645

```

$$C=18.81+0.99 * 1.645=18.97$$

So the rule of rejecting the null hypothesis when the sample mean is greater than or equal to 18.97 has the property that the probability of rejection is 5% when the null hypothesis is true.

## Analysis

We want to look at whether supplement types, dosages, or a combination of both had an effect on tooth growth. Let's first look at length by dosage and see if there is a correlation between the two.

```{r}
cor(ToothGrowth$len,ToothGrowth$dose)
```

The correlation is pretty close to 1, which means there is a positive correlation between the two -- the higher the dosage, the longer the length. Let's plot this with a linear regression line, which by default uses a 95% confidence region.

```{r message=FALSE, fig.height=3.75}
library(ggplot2)

ggplot(ToothGrowth, aes(dose, len)) +
  geom_point() +
  geom_smooth(method="lm") +
  ggtitle("Length of Teeth by Dosage per Day")

```


## State your conclusions and the assumptions needed for your conclusions. 